Satellite-detected droughts predict plant life history

### 2018-11-27 manuscript/manuscript.tex
version 3

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Satellite-detected drought frequencies predict plant life history strategies},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Satellite-detected drought frequencies predict plant life history
strategies}
    \author{J. Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3},
Kathryn Turner\textsuperscript{4}, \& John K. McKay\textsuperscript{2}}
    \date{}

@@

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of environmental stress such as drought. Here we test this
hypothesis in \textit{Heliophila} (Brassicaceae), a diverse genus of
flowering plants native to Southern Africa using herbarium occurrence
records and satellite-detected drought histories. We find that perennial
\textit{Heliophila} species are found in environments where droughts are
significantly less frequent compared to annuals. These associations are
predictive while controlling for phylogeny, lending support to the
hypothesis that drought related natural selection has influenced the
distributions of these strategies. Additionally, the difference in
drought frequency between annual and perennial species distributions is
greatest during the summer and fall, which also appears to be when
annuals are in the seed phase of their life cycle based on collection
dates of annual species. Together, these findings support traditional
hypotheses about the drivers of life history strategy in plants - that
perenniality is favored in environments with less frequent drought and
annuality in enviroments with more frequent drought by allowing them to
escape drought prone seasons as seeds. 
}

@@

\hypertarget{life-history-variation-in-plants}{%
\subsection{Life history variation in
plants}\label{life-history-variation-in-plants}}

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals (i.e.~monocarpic or semelparous)
which complete their seed to seed life cycle within a single calendar
year from perennials (i.e.~polycarpic or iteroparous{]} which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed. In contrast, perennial plants can continue vegetative growth after
reproduction and experience all seasons. These represent fundamentally
different life history strategies, but the ecological factors that
explain their evolution remain uresolved (Friedman \& Rubin, 2015).

Identifying the drivers of selection for these alternative strategies is
important because annuals and perennials have differing impacts on
ecosystem functioning. For example, perennials have higher nitrogen
concentration and larger specific root length in annuals, which both
affect nutrient cycling (Garnier, Cordonnier, Guillerm, \& Sonié, 1997;
Roumet, Urcelay, \& Dı'az, 2006). Furthermore, predicting the conditions
that favor these strategies in nature is useful for developing more
sustainable perennial cropping systems (Lelièvre \& Volaire, 2009).

Classical theory predict shorter life spans and annuality in
environments where adult mortality is high (Charnov \& Schaffer, 1973;
Franco \& Silvertown, 1996; Stearns, 1992). In plants, this has been
extended to the hypothessis that annuality is adaptive when it allows
plants to escape drought (Schaffer \& M, 1975). Lack of water is perhaps
the greatest threat to survival during vegetatitive or reproductive
growth and annuals can remain dormant (and protected as a seed) during
drought. Thus, environments with greater seasonal drought frequency may
select for annual life histories that complete reproduction prior to
drought prone seasons. Conversely, environments with less frequent
drought may select for perennial species, which may benefit from
multiple bouts of reproduction and competitive advantage (Corbin \&
D'Antonio, 2004). These prediction has been supported by the association
of annualilty with arid environments in wild rice (Morishima, Sano, \&
Oka, 1984) and \emph{Oenothera} (Evans, Hearn, Hahn, Spangle, \&
Venable, 2005). Additionally, annual and perennial species of Nemesia
were qualitatively associated with winter rather and summer rainfall
environments respectively (Datson, Murray, \& Steiner, 2008) and annual
species of Scorzoneroides were associated with environments classified
as unpredictable (Cruz-Mazo, Buide, Samuel, \& Narbona, 2009). These
findings provide qualitative evidence supporting the prediction that
annual species are found in environments that experience more frequent
drought, and vice versa, but whether the history frequency of drought
events indeed predicts the distributions annual or perennial life
history strategies has yet to be tested.

Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural collections to test these classic
hypotheses about the evolution of life history strategies within the
African endemic mustard genus, \emph{Heliophila} L. (Brassicaceae). If
annuallity is an adaptive strategy allowing plants to escape drought
prone seasons, then drought frequency should predict the distribution of
life history strategies across landscapes, and annual species should be
more commonly associated with drought prone regions than perennial
species. Furthermore, if annual species have adapted to escape drought
prone seasons as seeds, observations of annual species should be rare
during drought prone seasons. Phylogenetic relatedness can have
significant non-random effects on species distributions and life history
traits {[}{]}, and therefore we assessed the relationship between life
history distribution and drought frequency in a phylogenetically
controlled background.

@@

\hypertarget{availability}{%
\subsection{Availability}\label{availability}}

All data and the source code to produce this manuscript are available at
\url{https://github.com/greymonroe/heliophila}.

@@

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak et al., 2015). To quantify the frequency of drought during
different seasons across landscapes, we used the remotely sensed
Vegetative Health Index (VHI), which measures landscape scale reductions
in plant cover and temperature conditions characteristic of drought
(Kogan, 2001). Generated from data collected by NOAA AVHRR satellites
since 1981, the VHI combines normalized difference vegetation index
(NDVI) derived measures of vegetative stress (Vegetative Condition Index
- VCI) with temperature stress indicated by anomalies in thermal spectra
(Temperature Condition Index - TCI).

The VHI of year \(y\) during week \(w\) of \([1,52]\) at pixel \(i\) is
derived from the following equations, where \(n\) is the number of years
observed.

@@

\[VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})\]

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas, Vrieling, \& Rembold, 2011;
Kogan et al., 2016). But they also present a tool to study seasonal
patterns in the frequency of drought across environments and to test
hypotheses about the effect of drought on ecological and evolutionary
processes. Indeed, the VHI has been applied recently to study drought
related ecology of natural species and proven useful for predicting
infraspecific variation in drought tolerance traits and genes
(Dittberner et al., 2018; Mojica et al., 2016; Monroe et al., 2018).
Here, we accessed VHI data at \(16km^2\) resolution from 1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} sepecies.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the estimated \textasciitilde{}50 \emph{Heliophila} species
are considered to be the most diverse genus of the family Brassicaceae
(Mummenhoff, Al-Shehbaz, Bakker, Linder, \& Mühlhausen, 2005). This
genus includes both perennial and annual species and this change in life
history strategy has likely arisen multiple independent times (Appel \&
Al-Shehbaz, 1997; Mummenhoff et al., 2005). We used life histories
reported by Mummenhoff et al. (2005), grouping species with annual or
perennial life histories. Perenniality was defined based on the
traditional defition and included any form of perennial life history
(e.g., herbs, shrubs, mixed, etc).

\hypertarget{heliophila-herbarium-specimens}{%
\subsubsection{Heliophila herbarium
specimens}\label{heliophila-herbarium-specimens}}

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang, Willems, Scheepens, Burbano, and Bossdorf (2018)). And as they
become digitized (Soltis, 2017), these collections have been used to
study relationships between trait distributions, geography, and climate
(Davis, Willis, Connolly, Kelly, \& Ellison, 2015; Stropp et al., 2016;
Václavı'k, Beckmann, Cord, \& Bindewald, 2017; Wolf, Zimmerman,
Anderegg, Busby, \& Christensen, 2016). To characterize the diributions
of annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018.

@@

Aligned Heliophila ITS sequences were obtained from previous work by
Mandáková et al. (2012). Aethionema, Alliaria, Cardamine, Chamira, and
Rorippa ITS records from were downloaded from Genbank {[}{]} and aligned
together with Heliophila ITS sequences using MAFFT as outgroups {[}{]}.

@@

\hypertarget{herbarium-record-quality-control}{%
\subsubsection{Herbarium record quality
control}\label{herbarium-record-quality-control}}

To remove instance with spurious location data, we filtered raw GBIF by
restricting our analyses to include only:\\
* Species with reported life history\\
* Records with geospatial data\\
* Records without known geospatial issues\\
* Records from collection sites classified as land pixels\\
* Records from Africa\\
* Non-duplicate records (ie. identical species, location, collection
date)

@@

Model selection for construction of phylogeny was performed in
jModeltest2 with CIPRES (cite). Based on this analysis, \(GTR + L\) were
selected. Ultrametric phylogeny was estimated with branch lengths as
relative time (details).

\hypertarget{comparison-of-drought-frequency-between-annual-and-perennial}{%
\subsubsection{Comparison of drought frequency between annual and
perennial}\label{comparison-of-drought-frequency-between-annual-and-perennial}}

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. The relationship between
drought frequencies across each taxon's range and life habitat (annual
or perennial) was evaluated using Firth's penalized-likelihood logistic
regression and phylogenetic logistic regression.

@@

To test the hypothesis that annual and perennial plants reflect
adaptation to alternative drought environments we examined the landscape
distribution of life history strategies in the large and diverse mustard
genus, \emph{Heliophila} Figure \ref{fig:phylogeny}. Using both
herbarium specimen metadata and a 30 year dataset of satellite generated
climate information, we tested the prediction that annual species are
more often observed in drought-prone locations than perennial species,
when controlling for phylogenetic relatedness. We found that drought
frequency is significantly different between the distributions of annual
and perennial species, with annuals being found in environments with
significantly more frequent drought, and that this signal is strongest
during the summer. These results remain significant while controlling
for the phylogenetic relationships of Heliophila species, yielding
support for the role that natural selection has played in driving
contemporary distributions of these alternatives strategies in relation
to drought regimes.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/phylogeny} \caption{Phylogeny of Heliophila.}\label{fig:phylogeny}
\end{figure}

\hypertarget{gbif-records}{%
\subsubsection{GBIF records}\label{gbif-records}}

Out of 8670 \emph{Heliophila} GBIF recrods, 6634 were for species with
reported life history (Mummenhoff et al., 2005), 3653 had geospatial
data, 3627 did not have geospatial issues, 3460 were located on pixels
classified as land having drought measurements, 3457 were located in
Africa, 3162 were not duplicated. After all filtering steps, 2192
records for 42 species (Figure \ref{fig:mapsboxplots}, Table
\ref{tab:speciesmeanstable}) passed.

\hypertarget{drought-frequency}{%
\subsubsection{Drought frequency}\label{drought-frequency}}

Drought frequency was signicantly different between annual and perennial
species.

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of annual and perennial Heliophila. Drought
frequency during the winter, spring, summer and fall. Drought
frequencies during each season observed at the collection sites of
Heliophila records.}\label{fig:mapsboxplots}
\end{figure}

@@

\hypertarget{collection-dates-1}{%
\subsubsection{Collection dates}\label{collection-dates-1}}

@@

@@

\hypertarget{summary}{%
\subsection{Summary}\label{summary}}

We found that the distribution of annual and perennial species of
Heliophila is significantly predicted by satellite detected historic
drought frequencies. Annual species are found in environments that
experience more frequent drought during the summer and fall quarters
compared to perennials. This relationship was consistent while
controlling for phylogenetic relatedness among the taxa studied,
indicating that these distributions cannot be explained entirely by
common ancestry. These results support the hypothesis that natural
selection has played a role in shaping the contemporary distributions of
these alternative life-history strategies.

\hypertarget{relationship-to-previous-hypotheseswork}{%
\subsection{Relationship to previous
hypotheses/work}\label{relationship-to-previous-hypotheseswork}}

>>>>

\documentclass[man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Satellite-detected droughts predict plant life history},
            pdfauthor={J. Grey Monroe, Brian Gill, Kathryn Turner, \& John K. McKay},
            pdfkeywords={drought adaptation, life history evolution, remote sensing,
phylogeography, herbaria records},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

  \title{Satellite-detected droughts predict plant life history}
    \author{J. Grey Monroe\textsuperscript{1,2}, Brian Gill\textsuperscript{3},
Kathryn Turner\textsuperscript{4}, \& John K. McKay\textsuperscript{2}}
    \date{}

@@

\abstract{
Explaining variation in life history strategies is a long-standing goal
of evolutionary biology. For plants, annual and perennial life histories
are thought to reflect adaptation to environments that differ in the
frequency of environmental stress such as drought. Here we test this
hypothesis in \textit{Heliophila} (Brassicaceae), a diverse genus of
flowering plants native to Africa by integrating 2192 herbaria
occurrence records with 34 years of satellite-based drought detection.
Consistent with predictions from classic life history theory, we find
that perennial \textit{Heliophila} species occur in environments where
droughts are significantly less frequent compared to annuals. These
associations are predictive while controlling for phylogeny, lending
support to the hypothesis that drought related natural selection has
influenced the distributions of these strategies. Additionally, the
difference in drought frequency between annual and perennial species
distributions is greatest during the summer and fall, which also appears
to be when annuals are in the seed phase of their life cycle based on
collection dates of annual species. Together, these finding provide
empirical support for classic hypotheses about the drivers of life
history strategy in plants - that perennials out compete annuals in
environments with less frequent drought and that annuals are adpated to
enviroments with more frequent drought by escaping drought prone seasons
as seeds. 
}

@@

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals (i.e.~monocarpic or semelparous)
which complete their seed to seed life cycle within a single calendar
year from perennials (i.e.~polycarpic or iteroparous) which can persist
over multiple years. Annual plants flower once, set seed, senesce, and
then die, spending at least some portion of the year as a seed, where
they are relatively protected from environmental stress. In contrast,
perennial plants can continue vegetative growth after reproduction and
must survive conditions experienced during all seasons. These represent
fundamentally different life history strategies, but the ecological
factors that explain their evolution and distributions remain
empirically uresolved (Friedman \& Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Franco \& Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothessis that annuality is adaptive when it allows plants to escape
drought (Schaffer \& M, 1975). Lack of water is perhaps the greatest
threat to survival during vegetatitive or reproductive growth and
annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin \& D'Antonio, 2004). These predictions have been
supported by the association of annualilty with arid environments in
wild rice (Morishima, Sano, \& Oka, 1984) and \emph{Oenothera} (Evans,
Hearn, Hahn, Spangle, \& Venable, 2005). Additionally, annual and
perennial species of \emph{Nemesia} were qualitatively associated with
winter rather and summer rainfall environments respectively (Datson,
Murray, \& Steiner, 2008) and annual species of \emph{Scorzoneroides}
were associated with environments classified as unpredictable
(Cruz-Mazo, Buide, Samuel, \& Narbona, 2009). However, whether the
history frequency of drought events indeed predicts the distributions
annual or perennial life history strategies has yet to be tested in a
phylogenetic context with suitable sample sizes of distrubution and
climate data.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural collections to test these classic
hypotheses about the evolution of life history strategies within the
African endemic mustard genus, \emph{Heliophila} L. (Brassicaceae). If
annuallity is an adaptive strategy allowing plants to escape drought
prone seasons, then drought frequency should predict the distribution of
life history strategies across landscapes, and annual species should be
more commonly associated with drought prone regions than perennial
species. Furthermore, if annual species have adapted to escape drought
prone seasons as seeds, observations of annual species should be rare
during drought prone seasons. Phylogenetic relatedness can have
significant non-random effects on species distributions and life history
traits (Barrett, Harder, \& Worley, 1996), and therefore we assessed the
relationship between life history distribution and drought frequency in
a phylogenetically controlled background.

@@

@@

\hypertarget{availability}{%
\subsubsection{Availability}\label{availability}}

All analyses were peformed using R. All data and the source code to
produce this manuscript are available at
\url{https://github.com/greymonroe/heliophila}.

@@

Remotely sensed data is a powerful tool for characterizing seasonal
patterns in drought because it is less limited in spatial and temporal
scope and resolution than weather stations or field observations
(AghaKouchak et al., 2015). To quantify the frequency of drought during
different seasons across landscapes, we used the remotely sensed
Vegetative Health Index (VHI), which measures landscape scale reductions
in plant cover and temperature conditions characteristic of drought
(Kogan, 2001). Generated from data collected by NOAA AVHRR satellites
since 1981, the VHI combines normalized difference vegetation index
(NDVI) derived measures of vegetative stress (Vegetative Condition Index
- VCI) with temperature stress indicated by anomalies in thermal spectra
(Temperature Condition Index - TCI). The VHI of year \(y\) during week
\(w\) of \([1,52]\) at pixel \(i\) is derived from the following
equations, where \(n\) is the number of years observed.

@@

\[VHI_{y,w,i} = 0.5(VCI_{y,w,i}) + 0.5(TCI_{y,w,i})\] Thus, VHI
measurements are standardized according to conditions historically
observed at each locations. These measurements have been validated and
generally used for evaluating drought risk and predicting crop yields in
agriculture (e.g., Rojas, Vrieling, \& Rembold, 2011; Kogan et al.,
2016). But they also present a new tool to study seasonal patterns in
the frequency of drought across environments and to test hypotheses
about the effect of drought on ecological and evolutionary processes
(Kerr \& Ostrovsky, 2003). As such, the VHI has been applied recently to
study drought related ecology of natural species and proven useful for
predicting infraspecific variation in drought tolerance traits and genes
(Dittberner et al., 2018; Mojica et al., 2016; Monroe et al., 2018).
Here, we accessed VHI data at \(16km^2\) resolution from 1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the estimated \textasciitilde{}50 \emph{Heliophila} species
are considered to be the most diverse genus of the family Brassicaceae
(Mummenhoff, Al-Shehbaz, Bakker, Linder, \& Mühlhausen, 2005). This
genus includes both perennial and annual species and this change in life
history strategy has likely arisen multiple independent times (Appel \&
Al-Shehbaz, 1997; Mummenhoff et al., 2005). Furthermore, the fine scale
climatic heterogeneity of Southern Africa is ideal for studying the
distribution of traits in relation to environmental parameters (Sayre et
al., 2013). We used life histories reported by Mummenhoff et al. (2005),
grouping species with annual or perennial life histories. Perenniality
was defined based any form of perennial life history (e.g., herbs,
shrubs, mixed, etc). We also repeated these analyses without \emph{H.
subulata} and \emph{H. minima} which have mixed life histories and found
similar results. Because the original authors didn't report the nature
of the mixed trait (Mummenhoff et al., 2005), (ie. plasticity
vs.~genetic variation), we classified these species here as perennial
since they have the capacity to maintain vegetative growth after
reproduction at least to some capacity.

\hypertarget{heliophila-occurance-records}{%
\subsubsection{Heliophila occurance
records}\label{heliophila-occurance-records}}

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang, Willems, Scheepens, Burbano, and Bossdorf (2018)). And as they
become digitized (Soltis, 2017), these collections have been used to
study relationships between trait distributions, geography, and climate
(Davis, Willis, Connolly, Kelly, \& Ellison, 2015; Stropp et al., 2016;
Václavı'k, Beckmann, Cord, \& Bindewald, 2017; Wolf, Zimmerman,
Anderegg, Busby, \& Christensen, 2016). To characterize the diributions
of annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (``GBIF occurrence
download,'' 2018).

@@

Aligned Heliophila ITS sequences were obtained from previous work by
Mandáková et al. (2012). Aethionema, Alliaria, Cardamine, Chamira, and
Rorippa ITS records from were downloaded from Genbank.

@@

\hypertarget{filtering-of-occurance-records}{%
\subsubsection{Filtering of occurance
records}\label{filtering-of-occurance-records}}

To avoid instances with spurious location data, we filtered raw GBIF by
restricting our analyses to include only:

\begin{itemize}
\tightlist
\item
  Records for species with reported life history\\
\item
  Records with geospatial data\\
\item
  Records without known geospatial issues (i.e., coordinates reported
  are those of herbarium)\\
\item
  Records from collection sites classified as land pixels\\
\item
  Records from Africa\\
\item
  Records without duplicates (i.e., identical species, location,
  collection date)
\end{itemize}

@@

Out group species were aligned together with \emph{Heliophila} ITS
sequences using MAFFT. Model selection for construction of phylogeny was
performed in jModeltest2 with CIPRES. Based on this analysis,
\(GTR + L\) were selected. Ultrametric phylogeny was estimated with
branch lengths as relative time.

\hypertarget{comparison-of-drought-frequency-between-annual-and-perennial-species}{%
\subsubsection{Comparison of drought frequency between annual and
perennial
species}\label{comparison-of-drought-frequency-between-annual-and-perennial-species}}

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
raw occurance records of annual and perennial species by t-tests. To
account for variation in the number of occurance records per species, we
next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. The relationships between species mean
values of drought frequency during each season and life habitat (annual
or perennial) were tested using Firth's penalized-likelihood logistic
regressions. Because demographic histories caused by ancestry can
confound trait - environment associations, we then tested for the
relationships between drought frequeny and life history while
controlling for relationships between species using phylogenetic
logistic regressions.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/phylogeny} \caption{Species and examples of herbaria specimens of
\emph{Heliophila} (a) Phylogeny and life history strategies of species
studied. Orange circles at branch tips mark annual species and blue
circles mark perennial species. Example herbaria specimens accessed via
GBIF of (a) \emph{H. minima}, (b) \emph{H. deserticola}, (c) \emph{H.
coronopifolia} and (d) \emph{H. ephemera}. Images (a,c,d) courtesy of
The Bavarian Natural History Collections (CC BY-SA 4.0) and (b) The
London Natural History Museum (CC BY 4.0). Links to images are found in
the supplement.}\label{fig:phylogeny}
\end{figure}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/maps_boxplots} \caption{Locations of (a) annual and (b) perennial Heliophila.
Drought frequency during the (c) winter, (d) spring, (e) summer and (f)
fall. (g) Drought frequencies during each season observed at the
collection sites of \emph{Heliophila} records.}\label{fig:mapsboxplots}
\end{figure}

Out of 8670 \emph{Heliophila} GBIF recrods, 6634 were for species with
reported life history (Mummenhoff et al., 2005), 3653 had geospatial
data, 3627 did not have geospatial issues, 3460 were located on pixels
classified as land having drought measurements, 3457 were located in
Africa, 3162 were not duplicated. After all filtering steps, 2192
records for 42 species (Figure \ref{fig:phylogeny}, Table
\ref{tab:speciesmeanstable}) passed for further analyses. The number of
samples varied between species, with a mean of 52.19 samples per
species. Heliophila rigidiuscula had the most records, 201, and
Heliophila cornellsbergia the fewest, 2 (Table
\ref{tab:speciesmeanstable}).

There were clear visual differences between the distributions of the 960
annual and the 1232 perennial \emph{Heliophila} observation records (see
Figure \ref{fig:speciesmaps} for maps of individual species). While
annual species were generally found in the western regions of South
Africa and the Namibia, primarily in the Cape Floristic Region and
Succulent Karoo (Figure \ref{fig:mapsboxplots}a), the occurances of
perennials extended to the east coast of South Africa (Figure
\ref{fig:mapsboxplots}b).

The frequency of drought varied considerably across the ranges of
\emph{Heliophila} species (Figure \ref{fig:mapsboxplots}c-f). This
heterogeneity is expected, given that this is one of the most
climatically diverse regions of the Earth (Sayre et al., 2013). It is
worth noting the east to west cline in drought frequency observed during
the summer, which distinguishes the high drought frequency of the
Kalahari Sands and Namid Destert phyotogeographic regions from the low
drought frequency of the Drakensberg Mountains and Coastal Zambesian
phytogeographic regions. In the Cape phytogeographic region there was
finer scale heterogeneity in drought frequency during the summer.

Theory has predicted that annuality should be adaptive in places where
stresses such as drought are more common. Conversely, perenniality
should be adaptive in places where such stresses are less frequent. We
found that the frequency of drought was significantly higher at the
locations of occurance records for annual species. In terms of raw
observation records (Figure @\ref(fig:mapsboxplots)g), the frequency of
drought was signficantly higher at the location of annuals during the
winter (t = 10.65, p = 0.00), spring (t = 10.73, p = 0.00), summer (t =
12.67, p = 0.00), and fall (t = 15.26, p = 0.00). Because raw occurance
records do not account for variation in the number of records per
species (Table \ref{tab:speciesmeanstable})), we also tested whether
mean drought frequency values of each species were significantly
different between annuals and perennials by a Firths-penalized logistic
regression. We found that the mean drought frequencies were signficantly
higher (\(\alpha = 0.05\)) in annual species during the spring, summer,
and fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). We
further tested whether annual species are found in places where droughts
occur more frequently while controlling for the phylogentic relatedness
of \emph{Heliophila} species. This is important, because environmental
differences in species distributions can be confounded with demographic
history caused by acestry. We found that while controlling for
phylogeny, the mean drought frequencies were signficantly higher
(\(\alpha = 0.05\)) in annual species during the spring, summer, and
fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). These
findings indicate that common acestry alone does not explain differences
the drought frequencies experienced in the environemnts of annual and
perennial \emph{Heliophila}.

@@

@@

The preceding results indicated that annual speices are found in
environments where droughts are signifacantly more frequent, especially
in the summer and fall. Classic life history theory hypothesizes that
annuality reflects adaptation to such environments because it allows
species to escape stressful conditions. If this is the case, we would
expect that annuals spend the drought prone seasons of summer and fall
as seeds. To test this hypothesis, we compared the dates of occurance
records between annual and perennial \emph{Heliophila} species. The
distributions reveal a considerable difference in the timing of
observation of these two life histories. In comparison to perennials,
which appear to be collected throughout the year, annuals are almost
exclusively observed during the spring (Figure \ref{fig:lineplots}b).
The differences between the distrubution of collection dates were
signicant by all tests (ks.test D = 0.25, p = 0; bartlett.test K2 =
503.18, p = 0.00) This is consistent with a model of life history in
which annual species flower in the spring, set seed, senesce, and die
before the summer. Thus, these annual species are likely to remain
dormant during the summer and fall, when drought is the strongest
predictor of the distrubutions of annual and perennial life histories
(Figure \ref{fig:lineplots}a).

@@

To test the hypothesis that annual and perennial plants reflect
adaptation to alternative drought environments we examined the landscape
distribution of life history strategies in the large and diverse mustard
genus, \emph{Heliophila}. Using metadata of 2192 occurance records and a
34 year dataset of satellite-detected droughts, we tested the prediction
that annual species are more often observed in drought-prone locations
than perennial species, when controlling for phylogenetic relatedness.
We found that drought frequency is significantly different between the
distributions of annual and perennial species, with annuals being found
in environments with more frequent drought, and that this signal is
strongest during the summer. These results remain significant while
controlling for the phylogenetic relationships of \emph{Heliophila}
species, yielding support for the role that natural selection has played
in driving contemporary distributions of these alternatives strategies
in relation to drought regimes.
